PAMELA: an annotation-based Java Modeling Framework

### 2021-02-17 6-Experimentations.tex
remove orange comments (section 5.2 : Formose project)
experimentation after Antidote

 The PAMELA framework has been successfully applied in a variety of complex programming and modeling scenarios and we continue to use it daily as part of our modeling toolbox. In the following, we describe three important uses cases in which PAMELA was a core component.

@@

Model Federation~\cite{Golra2016} is an approach that provides the means to
integrate multiple models conforming to different paradigms, and giving to each
stakeholder a specific view adapted to its needs. Model federation approach is
developed as a possible response to SIMF RFP (Semantic Information Modeling for
Federation)~\cite{simf-rfp} by OMG (Object Modeling Group). This RFP (Request For Proposal)
requests submissions for a standard addressing, ”federation of information
across different representations, levels of abstraction, communities,
organizations, viewpoints, and authorities”. Thus model federation allows the
integration of heterogeneous models to develop new cross-concern
viewpoints/models or to synchronize the models used for designing a system.

Openflexo\cite{OpenflexoWebSite} is a software infrastructure providing support
for model federation across multiple technological spaces. Conceptualization is
addressed trough the proposition of a language called FML (Flexo Modeling
Language), which is executable on the platform. Openflexo infrastructure
introduces connectors (also called Technology Adapters) to support various
technological spaces and paradigms.

This open source initiative is now mature at the infrastructure level, and many
projects and applications have been developed and powered using Openflexo
infrastructure. More than 15 technology adapters have been developed, with
various maturity stage regarding their industrialization (Microsoft Word, Excel
and PowerPoint, EMF, OWL, Diagramming, JDBC, XML, OSLC, etc...)

The full Openflexo infrastructure is composed of about 50 components. In most components, PAMELA framework is extensively used. As an example \texttt{diana} component\footnote{\url{https://diana.openflexo.org}} (a component providing diagramming features) is composed of 998 classes. 159 of those classes (mostly the diagramming model) are defined as Pamela \texttt{ModelEntity}.

The total base of code for Openflexo infrastructure represents around 900.000
lines of Java code. Regarding backend modeling, the PAMELA framework is
extensively used in Openflexo core as well as in most technology adapters, with
very few specific implementations for properties. The observer/observable
pattern is generally used for graphical user interfaces, which also rely on
PAMELA framework.

An interesting experimentation has been done in the context of Openflexo development process. When Pamela has been integrated to the code base, a big portion of former legacy code has gradually and iteratively been migrated to Pamela. Refactoring mainly consisted in removing code, and replacing method implementation by API method declaration. In some parts of core model implementation, code has been reduced by 80\% (in terms of lines of Java code), and many bugs disappeared, as they were caused by programming errors.

Pamela implementation is now really stable and mature. According to Openflexo infrastructure developers, maintenance of code base (about 10 years of development) raises no Pamela-specific issue, while co-evolution of model and code is greatly improved compared to a "code generation"-based solution.

@@

The Formose ANR (french National Agency for Research) project
(ANR-14-CE28-0009)\cite{FormoseWebSite} aimed to design a formally-grounded,
model-based requirements engineering (RE) method for critical complex systems,
supported by an open-source environment. The main partners were: ClearSy, LACL,
Institut Mines-Telecom, OpenFlexo, and THALES.

The main results of the project are a requirements modeling multi-views
language, its associated design process and the development of an opensource
platform called Formod\cite{FormodWebSite}, built using Openflexo
infrastructure and PAMELA framework. \todo{Say how and where PAMELA was used}
The requirements modeling language is based on KAOS~\cite{kaos} for goal
modeling and SysML for the structural part of a system. The associated domain
modeling language, used to describe system domain knowledge, extends the two
ontology languages OWL~\cite{owl} and PLIB~\cite{plib}. The graphical notations
are then translated into Event-B~\cite{eventb}, a formal specification method supported by verification tools.

\todo{Remark that it is industry ready}

>>>>

 The PAMELA framework has been successfully applied in a variety of complex programming and modeling scenarios and we continue to use it daily as part of our modeling toolbox. In the following, we describe three important use cases in which PAMELA was a core component.

@@

Model Federation~\cite{Golra2016} is an approach that provides the means to
integrate multiple models conforming to different paradigms, and giving to each
stakeholder a specific view adapted to its needs. Model federation approach is
developed as a possible response to SIMF RFP (Semantic Information Modeling for
Federation)~\cite{simf-rfp} by OMG (Object Modeling Group). This RFP (Request For Proposal)
requests submissions for a standard addressing ”federation of information
across different representations, levels of abstraction, communities,
organizations, viewpoints, and authorities”. Thus model federation allows the
integration of heterogeneous models to develop new cross-concern
viewpoints/models or to synchronize the models used for designing a system.

Openflexo\cite{OpenflexoWebSite} is a software infrastructure providing support
for model federation across multiple technological spaces. Conceptualization is
addressed through the proposition of a language called FML (Flexo Modeling
Language), which is executable on the platform. Openflexo infrastructure
introduces connectors (also called Technology Adapters) to support various
technological spaces and paradigms.

This open source initiative is now mature at the infrastructure level, and many
projects and applications have been developed and powered using Openflexo
infrastructure. More than 15 technology adapters have been developed, with
various maturity stage regarding their industrialization (Microsoft Word, Excel
and PowerPoint, EMF, OWL, Diagraming, JDBC, XML, OSLC, etc.)

The full Openflexo infrastructure is composed of about 50 components. In most components, PAMELA framework is extensively used. As an example \texttt{diana} component\footnote{\url{https://diana.openflexo.org}} (a component providing diagraming features) is composed of 998 classes. 159 of those classes (mostly the diagraming model) are defined as Pamela \texttt{ModelEntity}.

The total base of code for Openflexo infrastructure represents around
\num{900000} lines of Java code. Regarding backend modeling, the PAMELA
framework is extensively used in Openflexo core as well as in most technology
adapters, with very few specific implementations for properties. The
observer/observable pattern is generally used for graphical user interfaces,
which also rely on PAMELA framework.

An interesting experimentation has been done in the context of Openflexo development process. When Pamela has been integrated to the code base, a big portion of the former legacy code has gradually and iteratively been migrated to Pamela. Refactoring mainly consisted in removing code, and replacing method implementation by API method declaration. In some parts of core model implementation, code has been reduced by 80\% (in terms of lines of Java code), and many bugs disappeared, as they were caused by programming errors.

Pamela implementation is now really stable and mature. According to Openflexo infrastructure developers, maintenance of code base (about 10 years of development) raises no Pamela-specific issue, while co-evolution of models and code is greatly improved compared to a "code generation"-based solution.

@@

The Formose ANR (French National Agency for Research) project
(ANR-14-CE28-0009)\cite{FormoseWebSite} aimed to design a formally-grounded,
model-based requirements engineering (RE) method for critical complex systems,
supported by an open-source environment. The main partners were: ClearSy, LACL,
Institut Mines-Telecom, OpenFlexo, and THALES.

The main results of the project are a requirements modeling multi-view
language, its associated design process and the development of an open-source
platform called Formod\cite{FormodWebSite}, built using Openflexo
infrastructure and PAMELA framework. %\todo{Say how and where PAMELA was used}
The requirements modeling language is based on KAOS~\cite{kaos} for goal
modeling and SysML for the structural part of a system. The associated domain
modeling language, used to describe system domain knowledge, extends the two
ontology languages OWL~\cite{owl} and PLIB~\cite{plib}. The graphical notations
are then translated into Event-B~\cite{eventb}, a formal specification method supported by verification tools.

%\todo{Remark that it is industry ready}
